---
title: "My title"
subtitle: "My subtitle"
author: 
  - Ziheng Zhong
thanks: "Code and data are available at: https://github.com/iJustinn/Toronto_Cycling_Network"
date: today
date-format: long
abstract: "My abstract"
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

```{r}
#| label: fig-one
#| fig-cap: one
#| echo: false



```



```{r}
#| label: fig-two
#| fig-cap: two
#| echo: false
#| warning: false
#| message: false



```



# Results

```{r}
#| echo: false
#| eval: true
#| label: tbl-one
#| tbl-cap: "cap"
#| warning: false



```

Our results are summarized in @tbl-modelresults.



# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.



\newpage



\appendix



# Appendix {-}



\newpage



# References


